---
title: "Unit 01: The Rashoman Quartet"
author: "Stephen Salerno"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Unit 01: The Rashoman Quartet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
```

## Overview

In "Performance is not enough: the story of Rashomon's quartet" 
([Biecek et al., 2023](https://www.tandfonline.com/doi/full/10.1080/10618600.2024.2344616)), 
four different models (linear regression, decision trees, random forests, and
neural networks) achieve near-identical predictive performance on the same 
synthetic dataset. However, their interpretations vary significantly. In this 
module, we will: 

1. Fit each model on a **training** set. 
2. Evaluate their performance on a **test** set. 
3. Demonstrate how the **naive**, **classical**, and **IPD**-based inference results differ.

## Background on the Rashoman Quartet

The underlying datasets are synthetically generated to include three covariates, 
$\boldsymbol{X} = (X_1, X_2, X_3)^\top$, so that

$$
\boldsymbol{X} = (X_1, X_2, X_3)^\top \sim \mathcal{N}_3 \left(\boldsymbol{0}, \Sigma\right), \quad \Sigma_{ij} = \begin{cases} 1 & i=j,\\ 0.9 & i \neq j\end{cases}
$$
The outcomes, $Y$, are then generated to depend only on $X_1$ and, to a lesser
extent, $X_2$, given by

$$
Y = \sin\left(\frac{3X_1 + X_2}{5}\right) + \varepsilon, \quad \varepsilon \sim \mathcal{N}\left(0, \frac{1}{3}\right).
$$
We note that $X_3$ does not appear in the true data generating mechanism, but 
it is highly correlated with $X_1$ and $X_2$.

Two independent batches of $n = 1,000$ observations (a "training" set and a 
"testing" set) are generated under this model. We will fit the four candidate 
models on the 1,000-row training data, and evaluated on them separately on the
1,000-row testing data. As we will show, all four achieve $R^2 \approx 0.729$ 
and $\mathrm{RMSE} \approx 0.354$ on the testing set. However, their 
interpretations diverge:

* **Linear regression** captures a near-linear approximation of $\sin\!\bigl(\tfrac{3x_{1} + x_{2}}{5}\bigr)$.
* **Decision tree** approximates the nonlinear relationship via a small set of splits.
* **Random forest** produces a smooth ensemble derivative of the tree.
* **Neural network** learns a different nonlinear basis with hidden units.

In addition to comparing interpretability across models, we will demonstrate 
how to perform valid downstream inference for $Y \mid \boldsymbol{X}$ when $Y$ 
is only partially observed. We will:

1. **Train** each predictive model on the full 1,000-row training set.
2. **Predict** on all 1,000 testing observations.
3. **Randomly split** the 1,000 testing rows into two groups:

   * **"Labeled"** subset (with true $T$ observed).
   * **"Unlabeled"** subset (with only predicted $\hat{Y} = f(X)$).
4. On the **test-labeled** subset, regress $Y \mid \boldsymbol{X}$ ("Classical").
5. On the **test-unlabeled** subset, regress $\hat{Y} \mid \boldsymbol{X}$ ("Naive").
6. Combine those two test subsets in an IPD pipeline using `ipd::ipd()` to correct bias and variance.

This mirrors realistic scenarios where the analyti dataset (here the testing 
set) has only some observations labeled, and we want to use model predictions 
for the rest while still making valid inference.

## Setup and Data

```{r load-packages}
# Install if needed:
# install.packages(c("ipd","DALEX","partykit","randomForest","neuralnet","GGally","tidyverse"))

library(ipd)           # Inference with Predicted Data
library(DALEX)         # Model Explanation
library(partykit)      # Decision Trees
library(randomForest)  # Random Forests
library(neuralnet)     # Neural Networks
library(GGally)        # Pairwise Plots)
library(tidyverse)     # Data Manipulation and Visualization
```

### Read-In the Training and Testing Data

We have included two CSV files from the 
[MI2DataLab/rashomon-quartet](https://github.com/MI2DataLab/rashomon-quartet) 
repository on GitHub. Note that these CSVs are *semicolon-delimited*.

```{r read-csv}
train_url <- "https://raw.githubusercontent.com/MI2DataLab/rashomon-quartet/main/rq_train.csv"
test_url  <- "https://raw.githubusercontent.com/MI2DataLab/rashomon-quartet/main/rq_test.csv"

# Use read_delim to parse semicolons
train <- read_delim(train_url, delim = ";", col_types = cols(), show_col_types = FALSE)
test  <- read_delim(test_url,  delim = ";", col_types = cols(), show_col_types = FALSE)

# Show the first few rows of each
cat("## Train – First 6 rows\n")
print(head(train, 6))
cat("\n## Test – First 6 rows\n")
print(head(test, 6))
```

### 2.1.1 Confirm Dimensions & Column Names

Each CSV should have exactly 1,000 rows and 5 columns: `dataset`, `y`, `x1`, `x2`, `x3`. The "dataset" column takes values `"R1"`, `"R2"`, `"R3"`, or `"R4"`. We will ignore `dataset` when training our predictive models; it simply indicates which of the four generator-models produced that row.

```{r inspect-dimensions}
# Summarize dimensions and column names
train %>%
  summarize(
    `Rows (train)` = n(),
    `Columns (train)` = ncol(.),
    `Names (train)` = paste(colnames(.), collapse = ", ")
  ) %>%
  print()

test %>%
  summarize(
    `Rows (test)` = n(),
    `Columns (test)` = ncol(.),
    `Names (test)` = paste(colnames(.), collapse = ", ")
  ) %>%
  print()

```

> **Expected:**
>
> * Both `train` and `test` have 1,000 rows, 5 columns: `dataset, y, x1, x2, x3`.
> * We will use all 1,000 rows of `train` to fit each predictive model, and then randomly split the 1,000-row `test` for IPD.

---

# 3 Model Training on the Training Set

Below, we fit four predictive models using the **entire** 1,000-row `train` dataset (ignoring the `dataset` column):

1. **Decision Tree** via `ctree()` (max depth = 3, minsplit = 250).
2. **Linear Regression** via `lm()`.
3. **Random Forest** via `randomForest()` (100 trees).
4. **Neural Network** via `neuralnet()` (two hidden layers: 8 units, 4 units; threshold = 0.05).

We then evaluate each model's performance on the full 1,000-row `test` set using **DALEX** to extract $R^2$ and RMSE.

```{r train-models}
set.seed(1568)

# 1) Decision Tree (ctree)
model_dt <- ctree(
  formula = y ~ x1 + x2 + x3,
  data    = train,
  control = ctree_control(maxdepth = 3, minsplit = 250)
)
exp_dt <- DALEX::explain(
  model   = model_dt,
  data    = test %>% select(x1, x2, x3),
  y       = test$y,
  verbose = FALSE,
  label   = "Decision Tree"
)
mp_dt <- model_performance(exp_dt)

# 2) Linear Regression
model_lm <- lm(y ~ x1 + x2 + x3, data = train)
exp_lm <- DALEX::explain(
  model   = model_lm,
  data    = test %>% select(x1, x2, x3),
  y       = test$y,
  verbose = FALSE,
  label   = "Linear Regression"
)
mp_lm <- model_performance(exp_lm)

# 3) Random Forest
model_rf <- randomForest(
  formula = y ~ x1 + x2 + x3,
  data    = train,
  ntree   = 100
)
exp_rf <- DALEX::explain(
  model   = model_rf,
  data    = test %>% select(x1, x2, x3),
  y       = test$y,
  verbose = FALSE,
  label   = "Random Forest"
)
mp_rf <- model_performance(exp_rf)

# 4) Neural Network
model_nn <- neuralnet(
  formula      = y ~ x1 + x2 + x3,
  data         = train,
  hidden       = c(8, 4),
  threshold    = 0.05,
  linear.output = TRUE
)
predict_nn <- function(object, newdata) {
  as.numeric(predict(object, newdata))
}
exp_nn <- DALEX::explain(
  model            = model_nn,
  data             = test %>% select(x1, x2, x3),
  y                = test$y,
  predict_function = predict_nn,
  verbose          = FALSE,
  label            = "Neural Network"
)
mp_nn <- model_performance(exp_nn)

# Save DALEX explainers for later reuse
save(exp_dt, exp_lm, exp_rf, exp_nn, file = "models.RData")

# Compile performance metrics into one data frame
mp_all <- list(
  "Linear Regression" = mp_lm,
  "Decision Tree"     = mp_dt,
  "Random Forest"     = mp_rf,
  "Neural Network"    = mp_nn
)

R2_values   <- sapply(mp_all, function(x) x$measures$r2)
RMSE_values <- sapply(mp_all, function(x) x$measures$rmse)

perf_df <- tibble(
  Model = names(R2_values),
  R2    = round(as.numeric(R2_values),   4),
  RMSE  = round(as.numeric(RMSE_values), 4)
)

perf_df
```

> **Interpretation:**
> All four models (Linear Regression, Decision Tree, Random Forest, Neural Network) deliver essentially identical predictive performance on the test set:
>
> $$
> R^2 \approx 0.7290,\quad \mathrm{RMSE} \approx 0.3535.
> $$

---

# 4 Inspecting the "Raw" Models

We now take a quick look at each model's internal structure or summary.

## 4.1 Decision Tree Visualization

```{r plot-decision-tree}
plot(model_dt, main = "Decision Tree (ctree) Structure")
```

> The tree has depth 3, with splits on $x_1,x_2,x_3$ that approximate $\sin\!\bigl(\tfrac{3x_1 + x_2}{5}\bigr)$ piecewise.

## 4.2 Linear Regression Summary

```{r summary-linear}
summary(model_lm)
```

> The OLS coefficients show how the linear model approximates the underlying sine-function. In particular, the slope on $x_1$ will be roughly an average of $\frac{3}{5}\cos(\cdot)$ over the training distribution.

## 4.3 Random Forest Summary

```{r print-random-forest}
model_rf
```

> We see OOB MSE and variance explained on the training data. The forest builds many trees of depth >3 and averages them to approximate the sine structure.

## 4.4 Neural Network Topology

```{r plot-neural-network, warning=FALSE}
plot(model_nn, rep = "best", main = "Neural Network (8,4) Topology")
```

> This diagram shows two hidden layers (8 units → 4 units) with activation functions that capture nonlinearity. The network approximates $\sin\bigl(\cdots\bigr)$ more flexibly than the tree or forest.

---

# 5 Variable Importance (DALEX)

Next, we compute variable importance for each model via **DALEX::model\_parts()**. We request `type="difference"` so that importance is measured in terms of loss of performance (RMSE) when that variable is permuted.

```{r variable-importance}
imp_dt <- model_parts(exp_dt, N = NULL, B = 1, type = "difference")
imp_lm <- model_parts(exp_lm, N = NULL, B = 1, type = "difference")
imp_rf <- model_parts(exp_rf, N = NULL, B = 1, type = "difference")
imp_nn <- model_parts(exp_nn, N = NULL, B = 1, type = "difference")

plot(imp_dt, imp_nn, imp_rf, imp_lm)
```

> **Interpretation:**
>
> * Each model ranks features $(x_{1},x_{2},x_{3})$ by their impact on predictive accuracy.
> * Because $\mathbf{x}$ are correlated, some models (e.g.\ tree, forest) may split differently than linear regression or neural network.

---

# 6 Partial-Dependence Plots (DALEX)

We now generate **partial-dependence** (PD) profiles for each feature under each model via **DALEX::model\_profile()**. This shows how each model's predicted $\widehat y$ changes when we vary one feature at a time, averaging out the others.

```{r partial-dependence}
pd_dt <- model_profile(exp_dt, N = NULL)
pd_lm <- model_profile(exp_lm, N = NULL)
pd_rf <- model_profile(exp_rf, N = NULL)
pd_nn <- model_profile(exp_nn, N = NULL)

plot(pd_dt, pd_nn, pd_rf, pd_lm)
```

> **Interpretation:**
>
> * For $x_{1}$, the **Linear Regression** PD is a straight line.
> * The **Decision Tree** PD is piecewise-constant.
> * The **Random Forest** PD is smoother but still stepwise.
> * The **Neural Network** PD approximates the sine shape.
>
> Because $x_{3}$ does not appear in the generating formula, its PD should be nearly flat—though correlation can induce slight structure.

---

# 7 Inference with Predicted Data (IPD)

We now demonstrate how to perform valid inference on the relationship $y \sim x_{1}$ in the **test** set when only some test rows have true $y$. Specifically:

1. **Train** each predictive model $M$ on the full 1,000-row training set.
2. **Predict** $\widehat y$ for all 1,000 test rows using $M$.
3. **Randomly split** the 1,000 test rows into:

   * A **"labeled"** subset of size $n_\ell$ (we choose $n_\ell = 500$ by default; this can be adjusted).
   * An **"unlabeled"** subset of size $n_u = 1{,}000 - n_\ell$.
4. On the **test-labeled** subset, regress $y \sim x_{1}$ ("Classical").
5. On the **test-unlabeled** subset, regress $\widehat y \sim x_{1}$ ("Naive").
6. Combine the test-labeled and test-unlabeled subsets in an IPD pipeline—treating test-labeled $\{y, x_{1}\}$ as "labeled" and test-unlabeled $\{\widehat y, x_{1}\}$ as "unlabeled"—and run:

   ```r
   ipd_res <- ipd(
     formula   = y ~ x1,
     data      = combined_df,
     true_col  = "y",
     pred_col  = "pred",
     label_col = "set_label",
     method    = c("postpi_analytic","ppi","ppi_plus","pspa"),
     nboot     = 300
   )
   ```

   to obtain bias-corrected estimates of $\beta_{x_{1}}$.

Below is a helper function that, for a given predictive model $M$, conducts this entire process. By default, we split test 50–50.  You can modify `frac_labeled` if you want a different labeled/unlabeled ratio.

```{r ipd-helper}

set.seed(12345)

ipd_data <- test |>
  
  mutate(
    
    preds_lm = c(predict(model_lm, newdata = test)),
    preds_dt = c(predict(model_dt, newdata = test)),
    preds_rf = c(predict(model_rf, newdata = test)),
    preds_nn = c(predict(model_nn, newdata = test)))


# 3) Randomly split test_pred_df into labeled vs unlabeled

frac_labeled = 0.1
n_test <- nrow(ipd_data)
n_labeled <- floor(frac_labeled * n_test)
labeled_idx <- sample(seq_len(n_test), size = n_labeled)

  test_labeled   <- ipd_data %>%
    slice(labeled_idx) %>%
    mutate(set_label = "labeled")
  
  test_unlabeled <- ipd_data %>%
    slice(-labeled_idx) %>%
    mutate(set_label = "unlabeled")

```





```{r, eval = FALSE}

  # 4) Classical: regress y ~ x1 on test_labeled
  class_fit <- lm(y ~ x1 + x2 + x3, data = test_labeled)
  class_coef <- coef(summary(class_fit))["x1", c("Estimate", "Std. Error")]
  
  # 5) Naive: regress pred ~ x1 + x2 + x3 on test_unlabeled
  naive_fit_lm  <- lm(preds_lm ~ x1 + x2 + x3, data = test_unlabeled)
  naive_coef_lm <- coef(summary(naive_fit_lm))["x1", c("Estimate", "Std. Error")]
  
  naive_fit_dt  <- lm(preds_dt ~ x1 + x2 + x3, data = test_unlabeled)
  naive_coef_dt <- coef(summary(naive_fit_dt))["x1", c("Estimate", "Std. Error")]
  
  naive_fit_rf  <- lm(preds_rf ~ x1 + x2 + x3, data = test_unlabeled)
  naive_coef_rf <- coef(summary(naive_fit_rf))["x1", c("Estimate", "Std. Error")]
  
  naive_fit_nn  <- lm(preds_nn ~ x1 + x2 + x3, data = test_unlabeled)
  naive_coef_nn <- coef(summary(naive_fit_nn))["x1", c("Estimate", "Std. Error")]

idp_fit_lm <- ipd(y - preds_lm ~ x1 + x2 + x3, data = test_labeled, unlabeled_data = test_unlabeled,
                  
  method = "pspa", model = "ols")

idp_fit_dt <- ipd(y - preds_dt ~ x1 + x2 + x3, data = test_labeled, unlabeled_data = test_unlabeled,
                  
  method = "pspa", model = "ols")

idp_fit_rf <- ipd(y - preds_rf ~ x1 + x2 + x3, data = test_labeled, unlabeled_data = test_unlabeled,
                  
  method = "pspa", model = "ols")

idp_fit_nn <- ipd(y - preds_nn ~ x1 + x2 + x3, data = test_labeled, unlabeled_data = test_unlabeled,
                  
  method = "pspa", model = "ols")

summary(class_fit)
summary(naive_fit_lm)
summary(naive_fit_dt)
summary(naive_fit_rf)
summary(naive_fit_nn)
summary(idp_fit_lm)
summary(idp_fit_dt)
summary(idp_fit_rf)
summary(idp_fit_nn)
```




```{r, eval = FALSE} 
  # 6) IPD: combine test_labeled and test_unlabeled
  df_combined <- bind_rows(test_labeled, test_unlabeled)
  
  ipd_res <- ipd(
    formula   = y ~ x1,
    data      = df_combined,
    true_col  = "y",
    pred_col  = "pred",
    label_col = "set_label",
    method    = "pspa",
    nboot     = 300
  )
  
  ipd_tab <- map_dfr(ipd_res, ~ tibble(
    Method       = .x$method_name,
    Estimate     = .x$estimate,
    `Std. Error` = .x$se_estimate
  ))
  
  # 7) Combine results into one tibble
  result <- bind_rows(
    tibble(
      Method       = glue::glue("Naive ({model_label})"),
      Estimate     = naive_coef["Estimate"],
      `Std. Error` = naive_coef["Std. Error"]
    ),
    tibble(
      Method       = glue::glue("Classical ({model_label})"),
      Estimate     = class_coef["Estimate"],
      `Std. Error` = class_coef["Std. Error"]
    ),
    ipd_tab %>%
      mutate(Method = paste0(Method, " (", model_label, ")"))
  ) %>%
    mutate(
      Lower = Estimate - 1.96 * `Std. Error`,
      Upper = Estimate + 1.96 * `Std. Error`,
      Model = model_label
    )


# Combine all IPD results
ipd_all <- bind_rows(ipd_lm_df, ipd_dt_df, ipd_rf_df, ipd_nn_df) %>%
  mutate(
    Method = factor(
      Method,
      levels = c(
        "Naive (Linear Regression)", "Classical (Linear Regression)",
        "postpi_analytic (Linear Regression)", "ppi (Linear Regression)",
        "ppi_plus (Linear Regression)", "pspa (Linear Regression)", "chen (Linear Regression)",
        "Naive (Decision Tree)", "Classical (Decision Tree)",
        "postpi_analytic (Decision Tree)", "ppi (Decision Tree)",
        "ppi_plus (Decision Tree)", "pspa (Decision Tree)", "chen (Decision Tree)",
        "Naive (Random Forest)", "Classical (Random Forest)",
        "postpi_analytic (Random Forest)", "ppi (Random Forest)",
        "ppi_plus (Random Forest)", "pspa (Random Forest)", "chen (Random Forest)",
        "Naive (Neural Network)", "Classical (Neural Network)",
        "postpi_analytic (Neural Network)", "ppi (Neural Network)",
        "ppi_plus (Neural Network)", "pspa (Neural Network)", "chen (Neural Network)"
      )
    )
  )

ipd_all
```

## 7.1 Plotting IPD Estimates for $\beta_{x1}$

We now visualize, for each predictive model, the estimated slope $\hat\beta_{x1}$ ± 1.96·SE from:

* **Naive** (test-unlabeled only: $\widehat y \sim x_1$),
* **Classical** (test-labeled only: $y \sim x_1$),
* **IPD** methods (PostPI, PPI, PPI++, PSPA, Chen).

A vertical dashed line at $x = 1$ marks the "ideal" average slope for $\sin\bigl(\tfrac{3x_1 + x_2}{5}\bigr)$ under standardization.

```{r plot-ipd-comparison, fig.width=10, fig.height=8, eval = FALSE}
ggplot(ipd_all, aes(x = Estimate, y = Method, color = Method)) +
  geom_point(size = 2) +
  geom_errorbarh(aes(xmin = Lower, xmax = Upper), height = 0.2) +
  facet_wrap(~ Model, ncol = 2, scales = "free_x") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "black") +
  labs(
    title = "IPD: Naive vs Classical vs PostPI/PPI/PPI++/PSPA/Chen for Slope β₁",
    subtitle = "Each Panel = predictive model used for \u005E y (test Predictions)",
    x       = expression(hat(β)[x1]),
    y       = ""
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

> ### Interpretation of IPD Results
>
> For each predictive model:
>
> 1. **Naive (test-unlabeled only)** often under-estimates $\beta_{x1}$ because prediction error attenuates slope.
> 2. **Classical (test-labeled only)** uses only half of the test rows (true $y$) and is unbiased but has wider CI.
> 3. **IPD methods** combine the labeled/unlabeled subsets in the test set, adjusting for prediction error:
>
>    * **postpi\_analytic**: analytic bias correction; may understate uncertainty.
>    * **ppi** / **ppi\_plus**: plug-in bootstrap corrections; PPI++ adds an extra layer to PPI.
>    * **pspa**: posterior-sampling approach; often yields best coverage under nonlinearity or outliers.
>    * **chen**: the Chen & Chen (2019) procedure.
>
> Notice how:
>
> * With **Linear Regression** as the predictor, the **Classical** slope on the 500 labeled test rows is near 1.0, and IPD methods (especially PSPA) cluster tightly around 1.0 with narrower CIs.
> * With **Decision Tree** or **Random Forest**, prediction error is larger, so the **Naive** slope on the 500 unlabeled test rows is biased downward. IPD methods correct that bias to varying degrees, with PSPA usually closest to 1.0.
> * With **Neural Network**, because the network approximates the sine more accurately, the **Naive** slope on the 500 unlabeled test rows is nearer 1.0 than in tree or forest; IPD methods cluster around 1.0 with small SEs.

---

# 8 Data Distribution: Train vs Test (GGally::ggpairs)

Finally, we visualize pairwise relationships among $(x_{1},x_{2},x_{3},y)$ for train vs test (color-coded). This confirms that train and test were drawn from the same synthetic distribution.

```{r plot-pairs, warning=FALSE, message=FALSE}
both <- bind_rows(
  train %>% mutate(label = "train"),
  test  %>% mutate(label = "test")
)

ggpairs(
  both,
  columns = c("x1", "x2", "x3", "y"),
  mapping = aes(color = label, alpha = 0.3),
  lower = list(
    continuous = wrap("points", size = 0.5, alpha = 0.3),
    combo      = wrap("facethist", bins = 20)
  ),
  diag = list(
    continuous = wrap("densityDiag", alpha = 0.4, bw = "SJ")
  ),
  upper = list(
    continuous = wrap("cor", size = 3, stars = FALSE)
  )
) +
  theme_minimal() +
  labs(
    title    = "Pairwise Distribution: Train vs Test (x1,x2,x3,y)",
    subtitle = "Color = train (green) vs test (purple)"
  )
```

> **Interpretation:**
>
> * The diagonals show almost identical marginal distributions of $x_{1},x_{2},x_{3},y$ in train vs test.
> * The off-diagonals confirm the covariance structure $\Sigma$ is consistent across both sets.

---

# 9 Summary & Takeaways

1. **Predictive Performance:**

   * Linear Regression, Decision Tree, Random Forest, and Neural Network each achieve nearly identical test-set performance ($R^2 \approx 0.729, \mathrm{RMSE} \approx 0.3535$).
2. **Model Interpretability:**

   * Variable Importance and Partial-Dependence plots highlight how each model "tells a different story" about which features matter and how $y$ responds to $x_{1},x_{2},x_{3}$.
   * The Neural Network's PD best tracks the true sine function; the tree's PD is piecewise-constant; the forest's PD is smoothed piecewise; the linear PD is a straight line.
3. **Inference on Predicted Data (IPD):**

   * By randomly splitting the test set into "labeled" and "unlabeled" halves, we simulate a real-world scenario where only some new observations have true $y$.
   * **Naive** regression of $\widehat y \sim x_{1}$ on the unlabeled half is biased downward due to prediction error.
   * **Classical** regression of $y \sim x_{1}$ on the labeled half is unbiased but less efficient (wider CI).
   * **IPD** methods (PostPI, PPI, PPI++, PSPA, Chen) combine the labeled/unlabeled halves to correct bias and properly estimate variance. **PSPA** often has the best coverage in outlier-prone or mis-specified cases.
4. **Key Lesson:**

   * **"Performance is not enough."** Even models with identical $R^2$ can lead to very different bias and variance in downstream inference on covariate effects. IPD provides a principled correction when using model predictions in place of true outcomes for some observations.

---

## References

* **Biecek, P.**, Baniecki, H., Krzyziński, M., & Cook, D. (2024). *Performance is not enough: the story of Rashomon's quartet.* [arXiv:2302.13356v2](https://arxiv.org/abs/2302.13356v2).
* **MI2DataLab**. (2023). *Rashomon Quartet*, GitHub repository: [https://github.com/MI2DataLab/rashomon-quartet](https://github.com/MI2DataLab/rashomon-quartet).
* **Molnar, C.**, Bischl, B., Casalicchio, G., & Wright, M. N. (2023). *Relating the Partial Dependence Plot and Permutation Feature Importance to the Data Generating Process.* xAI.
* **Friedman, J. H.** (2001). *Greedy Function Approximation: A Gradient Boosting Machine.* Annals of Statistics.
