---
title: "Unit 03: BCR-ABL Fusion in B-Cell Leukemia"
author: Stephen Salerno
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
rmarkdown::html_vignette:
  toc: true
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Unit 03: BCR-ABL Fusion in B-Cell Leukemia}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo     = TRUE,
  collapse = TRUE,
  message  = FALSE,
  warning  = FALSE,
  comment  = "#>"
)

# Load necessary packages
library(ipd)
library(ALL)            # ALL ExpressionSet
library(golubEsets)     # Golub_Merge ExpressionSet
library(AnnotationDbi)  
library(hgu95av2.db)    # Annotation for HGU95Av2
library(hu6800.db)      # Annotation for HGU6800
library(MLInterfaces)
library(pROC)
library(broom)
library(tidyverse)
```

## Background & Motivation

Acute lymphoblastic leukemia (ALL) is the most common childhood cancer, 
but it exhibits marked genetic heterogeneity, with distinct chromosomal 
translocations defining molecular subtypes with divergent prognoses and 
therapeutic responses. In B-cell lineage ALL, the **BCR-ABL1** fusion 
("breakpoint cluster region"/"Abelson", i.e., the "Philadelphia chromosome") 
arises from a t(9;22) translocation and encodes a constitutively active 
tyrosine kinase. This fusion was historically associated with poor outcomes, 
until the advent of targeted therapies (e.g., imatinib) revolutionized 
treatment and survival.

High-density microarray profiling measures expression of over 7,000 genes 
in leukemic blast cells, enabling a form of "molecular diagnostics" where 
supervised learning can:

1. **Classify** fusion status or other genetic subtypes when cytogenetic assays (PCR, FISH) are unavailable.
2. **Discover** novel marker genes and pathways dysregulated by specific fusions.
3. **Predict** therapeutic response and stratify risk using expression-derived scores.

However, in many retrospective or multi-center cohorts, only a subset of 
patients undergo gold-standard fusion testing (e.g., RT-PCR for BCR-ABL1), 
leaving the majority "unlabeled." A common workaround is to train a gene
expression classifier on the small labeled subset and apply it to the larger 
unlabeled cohort. Yet, naive downstream analyses, treating predicted labels 
as ground truth, can yield biased effect estimates and understate uncertainty.

[**Inference with Predicted Data (IPD)**](https://academic.oup.com/bioinformatics/article/41/2/btaf055/7997267) 
offers a principled remedy: it combines a small labeled set with the predicted 
labels (e.g., with true fusion status) with a larger unlabeled set, adjusting 
both bias and variance. In this module, we will:

1. **Describe** the [`ALL`](https://www.bioconductor.org/packages/release/data/experiment/html/ALL.html) and [`Golub`](https://www.bioconductor.org/packages/release/data/experiment/manuals/golubEsets/man/golubEsets.pdf) Bioconductor datasets.
2. **Subset** the `ALL` data to B-cell ALL and filter to the top 500 variable probes.
3. **Harmonize** probes across platforms via gene symbols.
4. **Split** the `ALL` data and train three classifiers for BCR-ABL1 fusion status.
5. **Predict** on the holdout `ALL` and `Golub` B-cell ALL data using the `ALL`-trained classifier.
6. **Apply** IPD to estimate the association between fusion status and patient sex assigned at birth.

By the end, you will understand how to leverage expression-based predictions 
while maintaining valid inference on genetic subtypes in cohorts with partially 
missing gold-standard labels. 

> **Note:** For the data loading and prediction workflows, we will directly 
> follow the Bioconductor MLInterfaces vignette by VJ Carey and P Atieno:
>
>
https://www.bioconductor.org/packages/devel/bioc/vignettes/MLInterfaces/inst/doc/MLprac2_2.html

## 1 Datasets Overview

We work with two public [Bioconductor](https://www.bioconductor.org/) datasets:

* [**ALL**](https://www.bioconductor.org/packages/release/data/experiment/html/ALL.html) (`ALL` package): 128 samples of acute lymphoblastic leukemia (both B- and T-cell), profiled on the Affymetrix HGU95Av2 microarray. Phenotype data include `BT` (immunophenotype), `mol.biol` (molecular subtype including BCR-ABL1 or NEG), age, and sex.
* [**Golub_Merge**](https://www.bioconductor.org/packages/release/data/experiment/manuals/golubEsets/man/golubEsets.pdf) (`golubEsets` package): 72 samples of leukemias (ALL vs. AML), profiled on Affymetrix HGU6800. Phenotype data include `ALL.AML`, `T.B.cell` (lineage), age, and sex.

We will train our classifiers on a subset of the `ALL` data, test the model on 
the holdout `ALL` data, and perform IPD on the holdout `ALL` and `Golub` data.

## 2 Phenotype Reduction and Feature Filtering

We first narrow to B-cell ALL samples and reduce dimensionality by keeping the 500 most variable probes.

```{r phenotype-filtering}
# Load the ALL data
data(ALL)

# Subset ALL to B-cell lineage (BT codes starting with "B")
bALL <- ALL[, substr(pData(ALL)$BT, 1, 1) == "B"]

# Keep only fusion-negative ("NEG") and fusion-positive ("BCR/ABL") samples
fus <- bALL[, bALL$mol.biol %in% c("NEG", "BCR/ABL")]

# Convert to factor with clear levels: NEG=0, BCR/ABL=1
fus$mol.biol <- factor(fus$mol.biol, levels = c("NEG", "BCR/ABL"))

# Compute median absolute deviation (MAD) for each probe
devs <- apply(exprs(fus), 1, mad)

# Select top 500 most variable probes by MAD ranking
top500 <- order(devs, decreasing = TRUE)[1:500]

# Subset ExpressionSet to top 500 probes
dat_filt <- fus[top500, ]

# Confirm dimensions: 500 probes x n samples
dim(exprs(dat_filt))
```

> We now have `dat_filt`, a filtered ALL dataset focused on B-cell lineage and the most informative probes.

## 3 Cross-Platform Probe Harmonization

To predict on the `Golub` data, which uses a different array, we map probes 
from each platform to gene symbols, then intersect to find a common gene set.

```{r harmonize}
# Load the Golub data
data(Golub_Merge)

# Map HGU95Av2 probes (ALL) to gene symbols
gmap_all <- AnnotationDbi::select(
  hgu95av2.db,
  keys    = featureNames(dat_filt),
  columns = c("PROBEID","SYMBOL"),
  keytype = "PROBEID"
)
# 3.2 Map HGU6800 probes (Golub) to gene symbols
gmap_golub <- AnnotationDbi::select(
  hu6800.db,
  keys    = featureNames(Golub_Merge),
  columns = c("PROBEID","SYMBOL"),
  keytype = "PROBEID"
)

# 3.3 Identify common symbols
common_sym <- intersect(gmap_all$SYMBOL, gmap_golub$SYMBOL)

# 3.4 For each symbol, pick the first associated probe
probe_all   <- gmap_all   %>% filter(SYMBOL %in% common_sym) %>% group_by(SYMBOL) %>% dplyr::slice(1) |> drop_na()
probe_golub <- gmap_golub %>% filter(SYMBOL %in% common_sym) %>% group_by(SYMBOL) %>% dplyr::slice(1) |> drop_na()

# 3.5 Subset ExpressionSets to these probes
dat_train  <- dat_filt[probe_all$PROBEID, ]
eset_golub <- Golub_Merge[probe_golub$PROBEID, ]

# 3.6 Further subset Golub to B-cell ALL samples
golub_bALL <- eset_golub[, pData(eset_golub)$T.B.cell == "B-cell" & pData(eset_golub)$ALL.AML == "ALL"]

featureNames(dat_train) <- probe_all$SYMBOL
rownames(exprs(dat_train)) <- probe_all$SYMBOL

featureNames(golub_bALL) <- probe_golub$SYMBOL
rownames(exprs(golub_bALL)) <- probe_golub$SYMBOL


# Confirm feature dimensions
dim(dat_train)    # features x ALL samples
dim(golub_bALL)   # features x Golub B-cell ALL samples
```

> `dat_train` and `golub_BALL` now share the same gene-symbol feature set, ready for model training and transfer.

---

## 4 Train/Test Split & Classifier Training

We split `dat_train` (ALL) into 60% training and 40% holdout, then train three ML models.

```{r train-test}
set.seed(2025)                # for reproducibility
n_all <- ncol(dat_train)      # number of ALL samples
train_idx <- sample(n_all, size = floor(0.6 * n_all))
# Training and holdout ExpressionSets
train_eset <- dat_train[, train_idx]
test_eset  <- dat_train[, -train_idx]
```

```{r train-classifiers}
# 4.1 Diagonal LDA (dldaI): assumes diagonal covariance for stability
dlda_mod <- MLearn(mol.biol ~ ., train_eset, dldaI, 1:40)
confuMat(dlda_mod)           # confusion matrix on training set

# 4.2 Neural Network (nnetI): single hidden layer size=5, decay=0.01
nn_mod <- MLearn(mol.biol ~ ., train_eset, nnetI, 1:40, size=5, decay=0.01, MaxNWts=2000)
confuMat(nn_mod)

# 4.3 Random Forest (randomForestI)
rf_mod <- MLearn(mol.biol ~ ., train_eset, randomForestI, 1:40)
confuMat(rf_mod) + confuMat(rf_mod, "train")
```

> Each confusion matrix shows true vs predicted counts on the **training** subset.  Accuracy around 75-85% reflects task difficulty.

---

## 5 Predictions on ALL Test and Golub

After fitting our classifiers on the **training** subset of ALL, we now generate predictions on both the **ALL holdout test** and the **Golub B-cell ALL** datasets.

```{r predict-all-test}
# 5.1 Prepare ALL test feature matrix
# test_mat_all <- t(exprs(test_eset)[probe_all$PROBEID, ])
# colnames(test_mat_all) <- probe_all$SYMBOL

# 5.2 Predict BCR-ABL1 status on ALL test holdout using random forest
all_pred <- MLInterfaces::predict.classifierOutput(rf_mod, test_eset)

# 5.3 Compare to true labels in test_eset
conf_mat_test <- table(
  Truth    = test_eset$mol.biol,
  Predicted= all_pred$testPredictions
)
conf_mat_test
```

```{r predict-golub}
# 5.4 Prepare Golub feature matrix (samples x genes)
# golub_mat <- t(exprs(golub_BALL)[probe_golub$PROBEID, ])
# colnames(golub_mat) <- probe_golub$SYMBOL

# 5.5 Predict BCR-ABL1 status on Golub B-cell ALL
golub_pred <- MLInterfaces::predict.classifierOutput(rf_mod, golub_bALL)

# 5.6 Summarize prediction counts (no ground truth in Golub)
table(golub_pred$testPredictions)
```

> **Note:** We retain **true** labels in the ALL test set for later IPD; Golub remains **unlabeled**.

---

## 6 IPD Inference on Sex Effect

We will now use **Inference with Predicted Data (IPD)** to estimate the log-odds effect of **sex** (male vs female) on true BCR-ABL1 status, combining the holdout ALL data with true and predict fusion status (i.e., our labeled data) and the Golub data with the predicted fusion status (i.e., our unlabeled data)

```{r ipd-setup}
# 6.1 Build labeled (ALL test) data frame
df_test <- tibble(
  sample = colnames(test_eset),
  set    = "labeled",
  true   = as.integer(test_eset$mol.biol == "BCR/ABL"),
  pred   = as.integer(all_pred$testPredictions == "BCR/ABL"),
  sex    = factor(pData(test_eset)$sex, levels = c("F","M"))
)

# 6.2 Build unlabeled (Golub) data frame
df_golub <- tibble(
  sample = colnames(golub_bALL),
  set    = "unlabeled",
  true   = NA_real_,
  pred   = as.integer(golub_pred$testPredictions == "BCR/ABL"),
  sex    = factor(pData(golub_bALL)$Gender, levels = c("F","M"))
)

# 6.3 Combine for IPD
ipd_df <- bind_rows(df_test, df_golub) |> drop_na(sex)
```

```{r ipd-run}
glm(true ~ sex, "binomial", df_test)

glm(pred ~ sex, "binomial", df_golub)

# 6.4 Run IPD logistic regression: true ~ sex
ires <- ipd(
  formula   = true - pred ~ sex,
  method    = "pspa",
  model     = "logistic",
  data      = ipd_df,
  label     = "set",
)

ires

# 6.5 Extract coefficient for sexM
df_ipd <- tidy(ires)
df_ipd
```

> **Insight:** By treating the ALL test set as labeled and Golub as unlabeled, IPD  combines information across cohorts to correct bias and improve precision when estimating the association between BCR-ABL1 status and patient sex.
