---
title: "Unit 00: Getting Started"
author: "Stephen Salerno"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    code_folding: "hide"
vignette: >
  %\VignetteIndexEntry{Unit 00: Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
```

## Overview

Welcome to this workshop on **Inference with Predicted Data (IPD)**! In this 
module, we will:

* Introduce the `ipd` package and its main functions.
* Demonstrate how to generate synthetic data for different models.
* Fit and compare various IPD methods.
* Inspect results using built-in `tidy`, `glance`, and `augment` methods.

Throughout the workshop, exercises are provided with solutions for practice.

## Background and Motivation

When an outcome $Y$ is costly or difficult to measure, it can be tempting to 
replace missing values with predictions $\hat{Y} = f(\boldsymbol{X})$ from a 
machine learning model (e.g., random forest or neural network) built on easier-to-measure features, $\boldsymbol{X}$. However, using $\hat{Y}$ as if 
it were the true outcome in downstream analyses, e.g., in estimating a 
regression coefficient, $\beta$, for $Y \mid \boldsymbol{X}$, leads to biased 
point estimates and underestimated uncertainty.  

Inference with predicted data (IPD) addresses this by collecting a small subset 
of "labeled" data with true $Y$ values to calibrate inference in a larger 
"unlabeled" set. The `ipd` package implements several recent methods for
IPD such as **Chen & Chen**, **PPI**, **PPI++**, **PostPI**, and **PSPA** to 
produce valid, efficient inference, even when a large proportion of outcomes 
are predicted.  

In this first tutorial, we demonstrate:

1. Generating fully synthetic data with `ipd::simdat()`.
2. Fitting a simple linear prediction model (e.g., linear regression).
3. Applying `ipd::ipd()` to estimate the slope, $\beta$, of $Y \mid X_1$ in the unlabeled set.
4. Comparing **naive**, **classical**, and **IPD** estimates of $\beta$.
5. Visualizing residuals and checking coverage.

## Installation and Setup

Ensure you have the `ipd` package and some additional packages installed:

```{r setup, class.source="fold-show"}
# Install these packages if you have not already:
# install.packages(c("ipd", "broom", "tidyverse", "patchwork"))

library(ipd)
library(broom)
library(tidyverse)
library(patchwork)
```

Throughout the workshop, we will use reproducible seeds and 
[tidyverse](https://www.tidyverse.org/) conventions.

## Function References

Below is a high-level summary of the core ``ipd`` functions.

### `simdat()`

Generates synthetic datasets for various inferential models.

```{r, class.source="fold-show", eval=FALSE}
simdat(
  n,       # Numeric vector of length 3: c(n_train, n_labeled, n_unlabeled)
  effect,  # Numeric: true effect size for simulation
  sigma_Y, # Numeric: residual standard deviation
  model    # Character: one of "mean", "quantile", "ols", "logistic", "poisson"
)
```
This function returns a data.frame with columns:

* `X1, X2, ...`: covariates
* `Y`: true outcome (for training, labeled, and unlabeled subsets)
* `f`: predictions from the model (for labeled and unlabeled subsets)
* `set_label`: character indicating "training", "labeled", or "unlabeled".

### `ipd()`

Fits IPD methods for downstream inference on predicted data.

```{r, class.source="fold-show", eval=FALSE}
ipd(
  formula,      # A formula: e.g., Y - f ~ X1 + X2 + ...
  method,       # Character: one of "chen", "postpi_boot", "postpi_analytic", 
                # "ppi", "ppi_all", "ppi_plusplus", "pspa"
  model,        # Character: one of "mean", "quantile", "ols", "logistic", 
                # "poisson"
  data,         # Data frame containing columns for formula and label
  label,        # Character: name of the column with set labels ("labeled" and 
                # "unlabeled")
  ...           # Additional arguments
)
```

#### Supported Methods

* **chen**: Chen and Chen estimator ([Gronsbell et al., 2025](https://arxiv.org/pdf/2411.19908))
* **postpi_analytic**: analytic post-prediction inference ([Wang et al., 2020](https://www.pnas.org/doi/10.1073/pnas.2001238117)).
* **postpi_boot**: bootstrap post-prediction inference [Wang et al., 2020](https://www.pnas.org/doi/10.1073/pnas.2001238117)).
* **ppi**: prediction-powered inference ([Angelopoulos et al., 2023](https://www.science.org/doi/10.1126/science.adi6000))
* **ppi_pluspus**: PPI++ (PPI with data-driven weighting; [Angelopoulos et al., 2024](https://arxiv.org/pdf/2311.01453))
* **ppi_a**: PPI using all data ([Gronsbell et al., 2025](https://arxiv.org/pdf/2411.19908))
* **pspa**: assumption-lean and data-adaptive post-prediction inference ([Miao et al., 2024](https://arxiv.org/pdf/2311.14220))

### Tidy Methods

* `print()` and `summary()`: display model summaries.
* `tidy()`: return a tibble of estimates and standard errors.
* `glance()`: return a one-row tibble of model-level metrics.
* `augment()`: return the original data with fitted values and residuals.

## Simulating Data

The `ipd::simdat()` function makes it easy to generate:

* A **training set** (where you fit your prediction model),
* A **labeled set** (where you observe the true $Y$),
* An **unlabeled set** (where $Y$ is *presumed missing* but you compute predictions $\hat{Y}$).

We supply `n = c(n_train, n_label, n_unlabel)`, an effect size (`effect`), 
residual standard deviation (`sigma_Y`), and a model type (`"ols"`, 
`"logistic"`, etc.). In this tutorial, we focus on a linear outcome. We can 
also optionally shift and scale the predictions (via the `shift` and `scale`
arguments) to control how the predicted outcomes relate to their true 
underlying counterparts.

### Exercise 1: Data Generation

Let us generate a synthetic dataset for an OLS model with:

* 5,000 training observations
* 500 labeled observations
* 1,500 unlabeled observations
* Effect size = 1.5
* Residual SD = 3
* Predictions shifted by 1 and scaled by 2

> **Note:** Interactive exercises are provided below. Try **writing your code** in the first chunk; **solutions** are hidden in the second chunk.

```{r, class.source="fold-show", eval=FALSE}
set.seed(123)
# Your code here
# dat <- simdat(...)
```

```{r}
set.seed(123)
# n_train = 5000, n_labeled = 500, n_unlabeled = 1500
n <- c(5000, 500, 1500)

# Effect size = 1.5, noise sd = 3, model = "ols" (ordinary least squares)
# We also shift the mean of the predictions by 1 and scale their values by 2
dat <- simdat(
  n       = n,
  effect  = 1.5,
  sigma_Y = 3,
  model   = "ols",
  shift   = 1,
  scale   = 2
)
```

```{r, class.source="fold-show"}
# The resulting data.frame `dat` has columns:
#  - X1, X2, X3, X4: Four simulated covariates (all numeric ~ N(0,1))
#  - Y             : True outcome (available in unlabeled set for simulation)
#  - f             : Predicted outcome (Generated internally in simdat)
#  - set_label     : {"training", "labeled", "unlabeled"}

# Quick look:
dat |> 
  group_by(set_label) |> 
  summarize(n = n()) 
```

Let us inspect the first few rows of each subset:

```{r, class.source="fold-show"}
head(dat[dat$set_label == "training", ])
head(dat[dat$set_label == "labeled", ])
head(dat[dat$set_label == "unlabeled", ])
```

> **Explanation:**
>
> * Rows where `set_label == "training"` form an internal training set.  Here, `Y` is observed, but `f` is `NA`, as we learn the prediction rule in this set.
> * Rows where `set_label == "labeled"` also have both `Y` and `f`.  In practice, 'f' will be generated by your own prediction model; for simulation, `simdat` does so automatically.
> * Rows where `set_label == "unlabeled"` have `Y` for posterity (but in a real‐data scenario, you would not know `Y`); `simdat` still places a placeholder in `Y`, but IPD routines will not use these.  The column `f` always contains 'predicted' values.

## Fit a Naive Linear Prediction Model

In a real application, we would take the **training** portion and fit an AI/ML 
model to predict $Y$ from `(X1, X2, X3, X4)`. This is done automatically by the
`simdat` function, but for demonstration, we will fit a 
*linear regression model* on the training data:

```{r, class.source="fold-show"}
# 1) Subset training set
dat_train <- dat |> 
  filter(set_label == "training")

# 2) Fit a linear model: Y ~ X1 + X2 + X3 + X4
lm_pred <- lm(Y ~ X1 + X2 + X3 + X4, data = dat_train)

# 3) Prepare a full-length vector of NA
dat$f_pred <- NA_real_

# 4) Identify the rows to predict (all non–training rows)
idx_analytic <- dat$set_label != "training"

# 5) Generate predictions just once on that subset (shifted and scaled to match)
pred_vals <- (predict(lm_pred, newdata = dat[idx_analytic, ]) - 1) / 2 

# 6) Insert them back into the full data frame
dat$f_pred[idx_analytic] <- pred_vals

# 7) Verify: `f_pred` is equal to `f` for the labeled and unlabeled data
dat |> 
  select(set_label, Y, f, f_pred) |> 
  filter(set_label != "training") |>
  head()
```

> **Explanation of arguments:**
>
> * `lm(Y ~ X1 + X2 + X3 + X4, data = dat_train)` fits an ordinary least squares (OLS) regression on the training subset.
> * `predict(lm_pred, newdata = .)` generates $\hat{Y}$ (stored as `f_pred`) for each row outside of the training set.

> **Note:** In real workflows, you might use random forests (`ranger::ranger()`), gradients (`xgboost::xgboost()`), or any other ML algorithm; the IPD methods only require that you supply a vector of predictions, `f`, in your data.

## Create 'Labeled' and 'Unlabeled' datasets

We now split the data into two subsets:

* **labeled**: those rows where we retain the true `Y` (to be used for final inference alongside their predictions).
* **unlabeled**: those rows where we **hide** the true `Y` (we pretend we do not observe them; `ipd` will still require the `f` + covariates).

```{r, class.source="fold-show"}
dat_ipd <- dat |>
  filter(set_label != "training") |>
  # Keep only the columns needed for downstream IPD
  select(set_label, Y, f, X1, X2, X3, X4) 

# Show counts:
dat_ipd |> 
  group_by(set_label) |> 
  summarize(n = n())
```

> **Explanation:**
>
> * After this step, `dat_ipd` has two groups:
>
>   * `labeled` (500 rows where we observe both `Y` and `f`),
>   * `unlabeled` (1500 rows where we only 'observe' `f`).

## Comparison of the True vs Predicted Outcomes

Before modeling, it is helpful to see graphically how the predicted values, 
$f$, compare to the true outcomes, $Y$. We can visually assess the **bias** 
and **variance** of our predicted outcomes, $f$, versus the true outcomes, 
$Y$, in our analytic data by plotting:

1. **Scatterplot** of $Y$ and $f$ vs. $X_1$  
2. **Density plots** of $Y$ and $f$  

```{r, class.source="fold-show"}
# Prepare data
dat_visualize <- dat_ipd |> 
  select(X1, Y, f) |>
  pivot_longer(Y:f, names_to = "Measure", values_to = "Value") |>
  arrange(Measure)

# Scatter + trend lines
ggplot(dat_visualize, aes(x = X1, y = Value, color = Measure)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c("steelblue", "gray")) +
  labs(
    x     = "X1",
    y     = "True Y or Predicted f",
    color = "Measure",
    title = "Scatterplot of True Outcome vs Prediction\nby Covariate X1"
  ) +
  theme_minimal()
```

```{r, class.source="fold-show"}
# Density plots
ggplot(dat_visualize, aes(x = Value, fill = Measure)) +
  geom_density(alpha = 0.4) +
  scale_fill_manual(values = c("steelblue", "gray")) +
  labs(
    x     = "Value",
    y     = "Density",
    fill  = "Measure",
    title = "Density of True Y vs Predicted f\n(showing reduced variance of predictions)"
  ) +
  theme_minimal()
```

> **Interpretation:**
>
> In the **scatterplot**, note that the predicted values $f$ (in blue) lie more tightly along the fitted trend line than the true $Y$ (in gray), indicating stronger correlation with $X_1$.
>
> In the **density plot**, you can see that the spread of $f$ is narrower than that of $Y$, illustrating that the predictive model has reduced variance (often due to "regression to the mean").

## Some Baselines: Naive vs Classical Inference

Before applying IPD, let's see what happens if we:

1. Regress *only the 'unlabeled'* predicted `f` on `X1` (the 'naive' approach).
2. Regress *only the 'labeled'* true `Y` on `X1` (the 'classical' approach).

We will compare these to IPD‐corrected estimates.

### Exercise 2: Naive vs. Classical Model Fitting

Using the `labeled` and `unlabeled` sets, fit two models:

1. **Naive OLS** on the unlabeled set using `lm()` with `f ~ X1`.
2. **Classical OLS** on the labeled set using `lm()` with `Y ~ X1`.

```{r, class.source="fold-show", eval=FALSE}
# Your code here
# naive_model <- lm(...)
# classical_model <- lm(...)
```

```{r}
# 1) Naive: treat f as if it were truth (only on unlabeled)
naive_model <- lm(f ~ X1, data = filter(dat_ipd, set_label == "unlabeled"))

# 2) Classical: regress true Y on X1, only on the labeled set
classical_model <- lm(Y ~ X1, data = filter(dat_ipd, set_label == "labeled"))
```

Let's also extract the coefficient summaries and compare the results

```{r, class.source="fold-show"}
naive_df <- tidy(naive_model) |>
  filter(term == "X1") |>
  select(estimate, std.error)

classical_df <- tidy(classical_model) |>
  filter(term == "X1") |>
  select(estimate, std.error)

tibble(
  Method       = c("Naive", "Classical"),
  Estimate     = c(naive_df$estimate,  classical_df$estimate),
  `Std. Error` = c(naive_df$std.error, classical_df$std.error)
)

```

> **Expected behavior (interpretation):**
>
> * The **naive** coefficient will typically be *attenuated* or *biased* toward zero if prediction is imperfect.
> * The **classical** coefficient (only 500 labeled pts) is unbiased but has large standard error due to smaller sample size.

## IPD: Corrected Inference with `ipd::ipd()`

The single wrapper function `ipd()` implements multiple IPD methods (e.g., 
**Chen & Chen**, **PostPI**, **PPI**, **PPI++**, **PSPA**) for various
inferential tasks (e.g., **mean** and **quantile** estimation, **ols**, 
**logistic**, and **poisson** regression under the hood.  

> **Basic usage of `ipd()`:**
>
> ```{r, class.source="fold-show", eval=FALSE} 
> ipd(
>   formula = Y - f ~ X1,     # The downstream inferential model
>   method  = "chen"          # The IPD method to run 
>   model   = "ols"           # The type of inferential model
>   data    = dat_ipd,        # A data.frame with columns:
>                             #   - set_label: {"labeled", "unlabeled"}
>                             #   - Y: true outcomes (for labeled data)
>                             #   - f: predicted outcomes
>                             #   - X covariates (here X1, X2, X3, X4)
>   label   = "set_label",    # Column name indicating "labeled"/"unlabeled"
> )
> ```

### Exercise 3: IPD Model Fitting via the Chen & Chen Estimator

Let us run one method, `chen` (the 'Chen & Chen' estimator proposed by 
[Gronsbell et al., 2025](https://arxiv.org/pdf/2411.19908)). The Chen & Chen 
estimator is an IPD method that combines information from:

1. **Labeled data** (where the true outcome, $Y$, and model predictions, $f$, are available), and  
2. **Unlabeled data** (where only model predictions, $f$, are available).

Rather than treating every unlabeled observation equally, it applies a 
**data-driven weight**, $\hat{W}^{\rm CC}$, to the unlabeled contribution and obtains

$$
\hat{\beta}^{\rm CC} = \hat{\beta}^{\rm lab} - \hat{W}^{\rm CC}(\hat{\gamma}^{\rm lab} - \hat{\gamma}^{\rm all})
$$

where $\hat{\beta}^{\rm lab}$ is the slope estimate from the classical 
regression, $\hat{\gamma}^{\rm lab}$ is the slope estimate from the naive
regression in the labeled data, $\hat{\gamma}^{\rm all}$ is the slope estimate
from the naive regression in the labeled + unlabeled data, and 
$\hat{W}^{\rm CC}$ reflects amound of additional information carried by the
predictions. By adaptively weighting the unlabeled information, the Chen & Chen estimator achieves greater precision than using labeled data alone, without 
sacrificing validity, even when the predictions are imperfect.  

Let's call the method using the `ipd()` function and collect the estimate for 
the slope of `X1` in a linear regression (`model = "ols"`):

```{r, class.source="fold-show", eval=FALSE}
set.seed(123)
# Your code here
# ipd_model <- ipd(...)
```

```{r}
set.seed(123)
ipd_model <- ipd(
  formula = Y - f ~ X1,
  data    = dat_ipd,
  label   = "set_label",
  method  = "chen",
  model   = "ols"
)

ipd_model
```

The `ipd_model` is an S4 object with slots for things like the coefficient, se, 
ci, coefTable, fit, formula, data_l, data_u, method, model, and intercept. We 
can extract the coefficient table using `ipd`'s `tidy` helper and compare with
the naive and classical methods:

```{r, class.source="fold-show"}
# Extract the coefficient estimates
ipd_df <- tidy(ipd_model) |>
  filter(term == "X1") |>
  select(estimate, std.error)

# Combine with naive & classical:
compare_tab <- tibble(
  Method       = c("Naive", "Classical", "IPD"),
  Estimate     = c(naive_df$estimate,  classical_df$estimate,  ipd_df$estimate),
  `Std. Error` = c(naive_df$std.error, classical_df$std.error, ipd_df$std.error)
)

compare_tab
```

### Exercise 4: Visualizing Uncertainty

Plot the coefficient estimates and 95% CIs for methods: naive, classical, and IPD.

```{r, class.source="fold-show", eval=FALSE}
# Your code: collect tidy outputs and plot
```

```{r, fig.width=6, fig.height=3}
# Forest plot of estimates and 95% confidence intervals
compare_tab |>
  mutate(
    Lower = Estimate - 1.96 * `Std. Error`,
    Upper = Estimate + 1.96 * `Std. Error`
  ) |>
  ggplot(aes(x = Estimate, y = Method)) +
    geom_point(size = 2) +
    geom_errorbarh(aes(xmin = Lower, xmax = Upper), height = 0.2) +
    geom_vline(xintercept = 1.5, linetype = "dashed", color = "red") +
    labs(
      title = "Comparison of slope estimates \u00B1 1.96·SE",
      x = expression(hat(beta)[X1]),
      y = ""
    ) +
    theme_minimal()
```

> **Interpretation:**
>
> * The dashed red line at **1.5** is the *true* data-generating effect for $X_1$.
> * Compare how far each method's interval is from 1.5, and whether 1.5 lies inside each interval.
> * 'Naive' often severely underestimates (biased); 'Classical' is unbiased but wide; IPD methods cluster around 1.5 with better coverage than 'naive,' often similar to classical but sometimes narrower.

### Exercise 5: Inspecting Results

Use `tidy()`, `glance()`, and `augment()` on `ipd_model`. Compare the 
coefficient estimate and standard error for `X1` with the naive fit.

```{r, class.source="fold-show", eval=FALSE}
# Your code here
```

```{r}
tidy(ipd_model)
glance(ipd_model)
head(augment(ipd_model))

# Compare with naive
broom::tidy(naive_model)
```

## Summary and Key Takeaways

1. **Naive** regression on predicted outcomes is biased (point estimates are pulled toward zero and SEs are artificially small).
2. **Classical** regression on the labeled data alone is unbiased but inefficient when the labeled set is small.
3. IPD methods (Chen & Chen, PPI, PPI++, PostPI, PSPA) strike a balance: they use predictions to effectively enlarge sample size but adjust for prediction error to avoid bias.
4. Even with 'simple' linear prediction models, IPD corrections can drastically improve inference on downstream regression coefficients.

## Further Exploration

* Try the **PPI++** (`"ppi_plusplus"`) and **PSPA** methods (`"pspa"`). How are they similar? How do they differ?
* Repeat the analysis for a **logistic** model by setting `model = "logistic"` in both `simdat()` and `ipd()`.

Happy coding! Feel free to modify and extend these exercises for your own data.

---

*This is the end of the module. We hope this was informative! For question/concerns/suggestions, please reach out to ssalerno@fredhutch.org*
